Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-09-30 source-iclr/sec_Nesterov.tex
Update on Overleaf.
iclr r1

\subsection{Strongly Convex and Smooth Objectives}
\begin{comment}
We first show that the Nesterov accelerated FedAvg has $\mathcal{O}(1/NT)$
convergence rate for $\mu$-strongly convex and $L$-smooth objectives.
\end{comment}
The Nesterov Accelerated FedAvg algorithm follows the updates:
\begin{align*}
\mathbf{v}_{t+1}^{k} & =\mathbf{w}_{t}^{k}-\alpha_{t}\mathbf{g}_{t,k}, \hspace{1em}
\mathbf{w}_{t+1}^{k} =\begin{cases}
\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k}) & \text{if }t+1\notin\mathcal{I}_{E},\\
\sum_{k \in \cS_{t+1}}q_k\left[\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k})\right] & \text{if }t+1\in\mathcal{I}_{E},
\end{cases}
\end{align*}
where $\mathbf{g}_{t,k}:=\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})$ is
the stochastic gradient sampled on the $k$-th device at time $t$, and $q_k$ again depends on participation and sampling schemes.  
\begin{theorem}
	\label{thm:nesterov_scvx}Let $\overline{\mathbf{v}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{v}_{T}^{k}$ in Nesterov accelerated FedAvg,
	and set learning rates $\alpha_{t}=\frac{6}{\mu}\frac{1}{t+\gamma}$,  $\beta_{t-1}=\frac{3}{14(t+\gamma)(1-\frac{6}{t+\gamma})\max\{\mu,1\}}$. Then under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right),
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right).
	\end{align*}
\end{theorem}
Similar to FedAvg, the key step in the proof of this result is a recursive contraction bound, but different in that it involves three time steps, due to the update format of Nesterov SGD:
\begin{align*}
\mathbb{E}\|\ov{v}_{t+1}-\vw^{\ast}\|^{2} & \leq\mathbb{E}(1-\mu\alpha_{t})(1+\beta_{t-1})^{2}\|\ov{v}_{t}-\vw^{\ast}\|^{2}+20E^{2}L\alpha_{t}^{3}G^{2}+(1-\alpha_{t}\mu)\beta_{t-1}^{2}\|(\ov{v}_{t-1}-\vw^{\ast})\|^{2}\\
& +\alpha_{t}^{2}\frac{1}{N}\nu_{\max}\sigma^{2}+2\beta_{t-1}(1+\beta_{t-1})(1-\alpha_{t}\mu)\|\ov{v}_{t}-\vw^{\ast}\|\cdot\|\ov{v}_{t-1}-\vw^{\ast}\|
\end{align*}
and from this one can again use induction and $L$-smoothness to obtain the desired bound.
%
To our knowledge, this is the first convergence result for Nesterov
accelerated FedAvg in the strongly convex and smooth setting. The
same discussion about linear speedup of FedAvg applies to the Nesterov
accelerated variant. In particular, to achieve $\mathcal{O}(1/NT)$
linear speedup, $T$ iterations of the algorithm require only $\mathcal{O}(\sqrt{NT})$
communication rounds with full participation.

@@

We now show that the optimality gap of Nesterov Accelerated FedAvg has $\mathcal{O}(1/\sqrt{KT})$ rate for convex and smooth objectives. This result complements the strongly convex case in the previous
part, as well as the non-convex smooth setting in \cite{huo2020faster,yu2019linear,li2018federated},
where a similar $\mathcal{O}(1/\sqrt{KT})$ rate is given in terms
of averaged gradient norm. 
\begin{theorem}
	\label{thm:Nesterov_cvx}Set learning rates $\alpha_{t}=\beta_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$. Then under Assumptions~\ref{ass:lsmooth},\ref{ass:boundedvariance},\ref{ass:subgrad2} Nesterov accelerated FedAvg with
	full device participation has rate
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T}\right),
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T}\right).
	\end{align*}
\end{theorem}
%
	It is possible to extend the results in this section to accelerated
    FedAvg algorithms with other momentum-based updates. However, as know from stochastic optimization, Nesterov and other momentum
updates may fail to accelerate over SGD (\cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}).
    For this reason, we will instead turn to the overparameterized setting
    \cite{ma2017power,liu2018accelerating,canziani2016analysis} in Section~\ref{sec:app:overparameterized} of the appendix where we show that FedAvg enjoys geometric convergence
    and it is possible to improve its convergence rate with a new momentum-based FedAvg variant, which we term ``FedMaSS''.

>>>>

\subsection{Strongly Convex and Smooth Objectives}
\begin{comment}
We first show that the Nesterov accelerated FedAvg has $\mathcal{O}(1/NT)$
convergence rate for $\mu$-strongly convex and $L$-smooth objectives.
\end{comment}
The Nesterov Accelerated FedAvg algorithm follows the updates:
\begin{align*}
\mathbf{v}_{t+1}^{k} & =\mathbf{w}_{t}^{k}-\alpha_{t}\mathbf{g}_{t,k}, \hspace{1em}
\mathbf{w}_{t+1}^{k} =\begin{cases}
\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k}) & \text{if }t+1\notin\mathcal{I}_{E},\\
\sum_{k \in \cS_{t+1}}q_k\left[\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k})\right] & \text{if }t+1\in\mathcal{I}_{E},
\end{cases}
\end{align*}
where $\mathbf{g}_{t,k}:=\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})$ is
the stochastic gradient sampled on the $k$-th device at time $t$, and $q_k$ again depends on participation and sampling schemes.  
\begin{theorem}
	\label{thm:nesterov_scvx}Let $\overline{\mathbf{v}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{v}_{T}^{k}$ in Nesterov accelerated FedAvg,
	and set learning rates $\alpha_{t}=\frac{6}{\mu}\frac{1}{t+\gamma}$,  $\beta_{t-1}=\frac{3}{14(t+\gamma)(1-\frac{6}{t+\gamma})\max\{\mu,1\}}$. Then under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	% \vspace{-1em}
	\small{\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right),
	\end{align*}}
	% \vspace{-1em}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\small{
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right).
	\end{align*}}
\end{theorem}
Similar to FedAvg, the key step in the proof of this result is a recursive contraction bound, but different in that it involves three time steps, due to the update format of Nesterov SGD (see Lemma~\ref{lem:nest-scvxoner} in Appendix~\ref{sec:convexsmoothsgd}). 
% \begin{align*}
% \mathbb{E}\|\ov{v}_{t+1}-\vw^{\ast}\|^{2} & \leq\mathbb{E}(1-\mu\alpha_{t})(1+\beta_{t-1})^{2}\|\ov{v}_{t}-\vw^{\ast}\|^{2}+20E^{2}L\alpha_{t}^{3}G^{2}+(1-\alpha_{t}\mu)\beta_{t-1}^{2}\|(\ov{v}_{t-1}-\vw^{\ast})\|^{2}\\
% & +\alpha_{t}^{2}\frac{1}{N}\nu_{\max}\sigma^{2}+2\beta_{t-1}(1+\beta_{t-1})(1-\alpha_{t}\mu)\|\ov{v}_{t}-\vw^{\ast}\|\cdot\|\ov{v}_{t-1}-\vw^{\ast}\|
% \end{align*}
Then we can again use induction and $L$-smoothness to obtain the desired bound.
%
To our knowledge, this is the first convergence result for Nesterov
accelerated FedAvg in the strongly convex and smooth setting. The
same discussion about linear speedup of FedAvg applies to the Nesterov
accelerated variant. In particular, to achieve $\mathcal{O}(1/NT)$
linear speedup, $T$ iterations of the algorithm require only $\mathcal{O}(\sqrt{NT})$
communication rounds with full participation.

@@

We now show that the optimality gap of Nesterov Accelerated FedAvg has $\mathcal{O}(1/\sqrt{KT})$ rate for convex and smooth objectives. This result complements the strongly convex case in the previous
part, as well as the non-convex smooth setting in \cite{huo2020faster,yu2019linear,li2018federated},
where a similar $\mathcal{O}(1/\sqrt{KT})$ rate is given in terms
of averaged gradient norm. 
\begin{theorem}
	\label{thm:Nesterov_cvx}Set learning rates $\alpha_{t}=\beta_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$. Then under Assumptions~\ref{ass:lsmooth},\ref{ass:boundedvariance},\ref{ass:subgrad2} Nesterov accelerated FedAvg with
	full device participation has rate
	% \vspace{-0em}
	\small{\begin{align*}
		\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T}\right),
		\end{align*}}
		% \vspace{-1em}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\small{\begin{align*}
		\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T}\right).
		\end{align*}}
\end{theorem}
%
	It is possible to extend the results in this section to accelerated
    FedAvg algorithms with other momentum-based updates. However, as know from stochastic optimization, Nesterov and other momentum
updates may fail to accelerate over SGD (\cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}).
    For this reason, we will instead turn to the overparameterized setting
    \cite{ma2017power,liu2018accelerating,canziani2016analysis} in Section~\ref{sec:app:overparameterized} of the appendix where we show that FedAvg enjoys geometric convergence
    and it is possible to improve its convergence rate with a new momentum-based FedAvg variant, which we term ``FedMaSS''.
